Factors affecting reliablity of state-space age-structured assessment models

### 2024-01-25 Ecov_study/mortality/paper/paper.tex

\hypertarget{summaryabstract}{%
\subsection*{Summary/Abstract}\label{summaryabstract}}
\addcontentsline{toc}{subsection}{Summary/Abstract}

We completed a large scale simulation study with 288 operating models
with 12 estimated models fit to each. The factors defining operating
model configuration inlcuded the source of process error on the
population (recruitment, recruitment and survival, recruitment and
natural mortality), the degree of temporal variation and autocorrelation
of the environmental covariate, the uncertainty in the observation of
the covariate, the uncertainty in indices and age composition
observations, fishing history, and the magnitude of the effect of the
covariate on natural mortality.

@@

AIC paragraph

Bias paragraph

These results suggest that models with covariate effects on M will be
most reliable when R+S is true, large contrast in the Recommendations
\#\#\# Keywords \{-\}

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each estimating model to the simulated data
sets. There are various ways to assess convergence of the fit. We
summarized 5 alternative categories of convergence:

\begin{enumerate}
\item Did the optimization routine (stats::nlminb) complete without error?
\item Did the stats::nlminb convergence flag = 0 indicate successful convergence?
\item Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
\item Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
\item Did TMB::sdreport provide all standard errors < 100?
\end{enumerate}

The first convergence criterion assesses whether the model crashes. The
third is a measure that assesses how flat the likelihood is at the
optimized point of the likelihood surface. The fourth and fifth criteria
are specific to the hessian-based standard error reporting provided by
TMB. the TMB::sdreport function will sometimes return standard error
estimates even when the calculated hessian is not invertible (fourth
criterion). It will also provide large standard errors for some
parameters that are not estimated well or are near bounds on the
transformed scale that is of primary interest. For example, variance
parameters for random effects can often be estimated near 0 when the fit
suggests no variation in the random effects or, equivalently, that the
model without random effects is better.

@@

We found high accuracy for estimation of the mean natural mortality rate
parameter (\(\beta_M\)) for all EM process errors assumptions when OMs
had step changes in fishing mortality, lower uncertainty in index and
age composition observations, and either R or R+M process errors (Figure
\ref{mean_M_bias}). The most variation in estimates occurred when
fishing and Higher uncertainty in index and age composition
observations. For OMs with R+S process errors, the most reliable
estimation of \(\beta_M\) was obtained when the EM also assumed R+S
process errors across all other factors. For OMs with R+M process
errors, the matching assumption for the EM only showed the best
reliability when there was low uncertainty in index and age composition
observations and fishing mortality was constant.

@@

We present results for error in annual natural mortality rate
conditional on three alternative EM configurations for the natural
mortality parameters: 1) mean natural mortality rate parameter is fixed
at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
\(\beta_\text{Ecov}\) are estimated.

@@

R+M EMs fit to R OMs with constant fishing mortality, or a step change
in fishing mortality and lower uncertainty in index and age composition
observations exhibited no bias in annual estimation of M indicating that
the variance of the estimated random effects for M in these EMs
collapsed to 0. R+M EMs were the only EMs to exhibit differences in the
sign of the median errors across the time series. When OMs had a step
change in fishing mortality and higher uncertainty in index and age
composition observations, we observed positive median errrors at the
beginning of the time series and lower median errors in later years.
However, whether there was evidence of bias, depended on the uncertainty
in covariate observations and the degree of variability in the latent
covariate. We observed negative median relative errors in annual M
estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and
age composition observations across all levels of simulated effects of
the covariate, but evidence of bias was strongest at the beginning of
the time series.

When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
OMs with evidence of negative bias for OMs with cosntant fishing
mortality, lower uncertainty in index and age composition observations,
the largest covariate effect size, lower variability in the latent
covariate, and larger uncertainty in the covariate observations (Figure
\ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference
in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated
or not. The results for R+M EMs were generally similar to those when
\(\beta_\text{Ecov}=0\) was assumed, except that the differences between
the estimates at the beginning of the time series and later years for
certain OM configurations did not occur.

Allowing the EMs to estimate \(\beta_M\) resulted in very large
variability in estimates for all EMs for R and R+M OMs with constant
fishing mortality and higher uncertainty in index and age composition
observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
The same varibilities occurred for R+S OMs, but strong bias was
estimated for R and R+M EMs. Annual M estimation was most reliable when
OMs had step changes in fishing mortality and lower uncertainty in index
and age composition observations. For R and R+M OMs with those
characteristics, all EMs generally provided accurate estimation of
annual M, but only R+S EMs provided accurate estimation for R+S OMs.
Little or no bias in annual M estimation was observed for R+S EMs across
all OM process error assumptions as long as there was a step change in
fishing mortality and lower uncertainty in indices and age composition.

>>>>

\hypertarget{summary}{%
\subsection*{Summary}\label{summary}}
\addcontentsline{toc}{subsection}{Summary}

We completed a large scale simulation study with 288 operating models,
each with 100 simulated data sets, and 12 estimated models fit to each
simulated data set. The factors defining operating model configuration
included the source of process error on the population (recruitment,
recruitment and survival, recruitment and natural mortality), the degree
of temporal variation and autocorrelation of the environmental
covariate, the uncertainty in the observation of the covariate, the
uncertainty in indices and age composition observations, fishing
history, and the magnitude of the effect of the covariate on natural
mortality. The estimating models make alternative assumptions on whether
to include the environmental effect, whether the mean/intercept log
natural mortality (log(0.2)) is estimated or known, and whether process
errors are on just recruitment, recruitment and survival, or recruitment
and natural mortality.

@@

Whether the mean log-natural mortality was estimated or not, the best
accuracy of AIC for model selection occurred for models with process
errors on recruitment and survival. AIC accuracy was poor for models
with process errors on recruitment and natural mortality. estimating the
mean natural mortality rate had small effects on the accuracy of AIC in
selecting the appropriate process error. Estimating the mean log-natural
mortality resulted in a small decrease in AIC accuracy. AIC was
conservative for determining whether the environmental covariate
affected natural mortality. AIC was very accurate in determining no
effect when there was no effect in the operating model, but AIC often
ranked the null model best when there was an effect. Accuracy of AIC for
covariate effects improved with increased effect size, increased
temporal contrast in the covariate, and lower uncertainty in
observations.

We found no evidence of bias in estimation of environmental effects
regardless of process error assumptions when there was low uncertainty
in the environmental observations and large constrast in the
environmental covariate. In most cases the relative error of the
estimated environmental effect did not depend on the source of process
error assumed in the estimating model. The worst bias was observed when
OMs assumed R+S process errors, high uncertainty in covariate
observations, low variability in the covariate, and low uncertainty in
index and age composition observations. Simultaneously estimating the
mean/intercept log natural mortality resulted in larger variation in the
relative errors of the estimated environmental effect. Estimation of the
intercept was reliable for all EM process error assumptions when the
operating models assumed process errors on recruitment and natural
mortality, contrast in fishing pressure over time, and lower observation
error. Estimating the mean/intercept for log natural mortality generally
resulted in highly variable estimates of annual natural mortality and
spawning biomass and evidence of bias for some operating and estimation
model assumptions about process error source. Again reliability of
annual natural mortality estimates was generally improved with lower
observation error uncertainty and contrast in fishing pressure.

Reliable detection of covariate effects requires informative data. AIC
preferred simpler models than the true model when information content in
data and contrast in covariates and abundance were low. The null model
for environmental covariate effects (no covariate effect) was selected
when contrast in the time series was low and/or uncertainty in
observations was high. The selection of the null model by AIC also
likely decreases with strength of the effect of the covariate on M.
Similarly, when there was process error in recruitment and natural
mortality, estimation models with process error only in recruitment were
preferred presumably due to low variation in simulated natural mortality
process errors. Covariate effect estimation can be robust to process
error assumptions with high contrast in covariate and low observation
error.

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each estimating model to the simulated data
sets. There are various ways to assess convergence of the fit, but we
defined successful convergence as the hessian of the marginal
log-likelihood being invertible and providing variance estimates for the
fixed effects parameters.

@@

We found high accuracy for estimation of the mean natural mortality rate
parameter (\(\beta_M\)) for all EM process errors assumptions when OMs
had step changes in fishing mortality, lower uncertainty in index and
age composition observations, and either R or R+M process errors (Figure
\ref{mean_M_bias_Ecov_beta_fixed}). The most variation in estimates
occurred when fishing mortality was constant and there was higher
uncertainty in index and age composition observations. For OMs with R+S
process errors, the most reliable estimation of \(\beta_M\) was obtained
when the EM also assumed R+S process errors across all other factors.
For OMs with R+M process errors, the matching assumption for the EM only
showed the best reliability when there was low uncertainty in index and
age composition observations and fishing mortality was constant.

@@

We present results for error in annual natural mortality rate
conditional on three alternative EM configurations for the natural
mortality parameters: 1) mean natural mortality rate parameter is fixed
at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
assumed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
\(\beta_\text{Ecov}\) are estimated.

@@

R+M EMs fit to R OMs with constant fishing mortality, or a step change
in fishing mortality and lower uncertainty in index and age composition
observations exhibited no bias in annual estimation of M indicating that
the variance of the estimated random effects for M in these EMs
collapsed to 0. R+M EMs were the only EMs to exhibit differences in the
sign of the median errors across the time series. When OMs had a step
change in fishing mortality and higher uncertainty in index and age
composition observations, we observed positive median errors at the
beginning of the time series and lower median errors in later years.
However, whether there was evidence of bias, depended on the uncertainty
in covariate observations and the degree of variability in the latent
covariate. We observed negative median relative errors in annual M
estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and
age composition observations across all levels of simulated effects of
the covariate, but evidence of bias was strongest at the beginning of
the time series.

When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
OMs with evidence of negative bias for OMs with constant fishing
mortality, lower uncertainty in index and age composition observations,
the largest covariate effect size, lower variability in the latent
covariate, and larger uncertainty in the covariate observations (Figure
\ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference
in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated
or not. The results for R+M EMs were generally similar to those when
\(\beta_\text{Ecov}=0\) was assumed, except that the differences between
the estimates at the beginning of the time series and later years for
certain OM configurations did not occur.

Allowing the EMs to estimate \(\beta_M\) resulted in very large
variability in estimates for all EMs for R and R+M OMs with constant
fishing mortality and higher uncertainty in index and age composition
observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
The same variability occurred for R+S OMs, but strong bias was estimated
for R and R+M EMs. Annual M estimation was most reliable when OMs had
step changes in fishing mortality and lower uncertainty in index and age
composition observations. For R and R+M OMs with those characteristics,
all EMs generally provided accurate estimation of annual M, but only R+S
EMs provided accurate estimation for R+S OMs. Little or no bias in
annual M estimation was observed for R+S EMs across all OM process error
assumptions as long as there was a step change in fishing mortality and
lower uncertainty in indices and age composition.

### 2024-01-25 Project_0/paper/paper.tex

We used the Woods Hole Assessment Model (WHAM) to configure operating
and estimation models in our simulation study \citep[]{stockmiller21}.
WHAM is an R package freely available as a github repository. For this
study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This packages has also been used to
configure operating and estimating models for closed loop simulations
evaluating index-based assessment methods \citep{legaultetal23} and is
used for management of haddock, butterfish, American plaice, bluefish,
Atlantic cod, and black sea bass in the Northeast US.

@@

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models \citep{millerbrooks21}. We specified unfished
recruitment = \(R_0 = e^{10}\) and \(\Fmsy = \Fspr[40] = 0.348\) equated
to a steepness of 0.69 and \(\alpha=0.60\) and
\(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean rate was 0.2.

>>>>

We used the Woods Hole Assessment Model (WHAM) to configure operating
and estimation models in our simulation study
\citep[]{millerstock20,stockmiller21}. WHAM is an R package freely
available as a github repository. For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This packages has also been used to
configure operating and estimating models for closed loop simulations
evaluating index-based assessment methods \citep{legaultetal23} and is
used for management of haddock, butterfish, American plaice, bluefish,
Atlantic cod, and black sea bass in the Northeast US.

@@

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models \citep{millerbrooks21}. We specified unfished
recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated
to a steepness of 0.69 and \(\alpha=0.60\) and
\(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean rate was 0.2.
